The importance of transient resonances in extreme-mass-ratio inspirals

### 2016-08-25 res_char.tex
included some suggestions

@@

\begin{abstract}
The inspiral of stellar-mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect and analyse these signals, accurate waveform templates are required. For computational efficiency, adiabatic templates are often used, which accurately reproduce orbit-averaged trajectories arising from the first-order dissipative gravitational self-force. Other effects are neglected, in particular that of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact object's trajectory and, ultimately, the phase of the emitted radiation. We investigate the properties of transient resonances. We present an evolution scheme for studying the effects of resonances and apply this to an astrophysically motivated population. We find that a large proportion encounter a low-order resonance in the later stages of inspiral; however, the resulting effect on signal-to-noise recovery is small as a consequence of the low eccentricity of the inspirals. Neglecting the effects of transient resonances leads to a loss of $4\%$ of detectable signals.
\end{abstract}

@@

\section{Introduction}

@@

We have made progress in understanding the general relativistic two-body problem in recent years. Bodies of comparable mass can be studied using numerical relativity. Rapid advances in this field have been made following breakthroughs in 2005~\cite{Pretorius2005,Campanelli2006,Baker2006}; it is now possible to simulate hundreds of orbits~\cite{Szilagyi2015}. These simulations allow us to understand binary BH coalescences. Stellar-mass BH mergers are targets for ground-based GW detectors, such as Advanced LIGO~\cite{Aasi2015} and Advanced Virgo~\cite{Acernese2015}, and the in-construction KAGRA~\cite{Aso2013}. The first direct observations of GWs came from the coalescences of two stellar-mass BHs~\cite{Abbott2016,Abbott2016e,Abbott2016d}, and analysis of their properties relied upon our understanding of binary BH waveforms~\cite{Abbott2016f,Abbott2016h,Abbott2016d}. %Supermassive BH (SMBH) coalescences, expected to be the result of galaxy mergers, are potential sources for a space-borne detector such as eLISA~\cite{Amaro-Seoane2012a}.

@@

In the case of extreme-mass-ratio systems, efforts are concentrated on understanding the gravitational self-force~\cite{Barack2009,Poisson2004}. In the test-particle limit, the smaller body follows an exact geodesic of the SMBH's spacetime. Including the effects of the smaller body's finite mass, the background spacetime is perturbed. The back-reaction from this deformation alters the small body's orbital trajectory, and can be modelled as a self-force that moves the body from its geodesic. The self-force is commonly divided into two pieces, dissipative and conservative. The former encapsulates the slow decay of the orbital energy and angular momentum, constants of the motion in the test particle limit, through radiation. The latter shifts the orbital phases inducing precession. The dissipative piece is time-asymmetric and has the larger effect on the evolution of the orbital phase; the conservative piece is time-symmetric and has a smaller influence on the phase, although this can accumulate over many orbits. Being able to accurately model the influence of the self-force allows us to create reliable waveform models.

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$.\footnote{In the strong-field regime, it is possible to have isofrequency pairings, where two different orbits share the same orbital frequencies~\cite{Warburton2013}. The evolution of the frequencies still differ, such that orbital trajectories can be reconstructed.} The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field regime $\Omega_r < \Omega_\theta < \Omega_\phi$, and they may differ significantly. For EMRIs, the evolution timescale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital timescales. The evolution of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate. Resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu \equiv \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

Resonances involving the azimuthal motion do not produce a comparable effect because of the axisymmetry of the background spacetime. However, both $\theta$--$\phi$ resonances~\cite{Hirata2011} and $r$--$\phi$ resonances~\cite{VanDeMeent2013} can lead to extrinsic effects; the GWs from such systems are not emitted isotropically and the imbalance produces a kick velocity that is, in some cases, sufficient to eject the central BH from its host.

Geodesic motion in Kerr spacetime can be described by use of the action--angle formalism~\cite{Goldstein2002}. % Chapter 10
We consider a body of mass $\mu$ orbiting a BH of mass $M$, with $\eta = \mu/M \ll 1$,\footnote{To first order, the mass ratio $\eta$ is the same as the symmetric mass ratio $\mu M/(\mu+M)^2$.} and describe the motion in the directions of the standard Boyer--Lindquist coordinates $\{t,r,\theta,\phi\}$ using generalised angle variables $q_\alpha = \{q_t,q_r,q_\theta,q_\phi\}$ \citep{Hinderer2008}. We denote the first integrals of the geodesic motion, the generalised action variables, by $J_\alpha$. These are some combination of the energy per unit mass $E$ and the axial angular momentum per unit mass $L_z$, which arise from isometries of the metric in $t$ and $\phi$, and the Carter constant per unit mass squared $Q$~\cite{Carter1968}, which is related to the separability of the equations of motion in $r$ and $\theta$. The system evolves following~\cite{Flanagan2012}
\begin{subequations}
\label{eq:Mino-E-o-M}
\begin{align}
\diff{q_\alpha}{\lambda} = {} & \omega_\alpha(\boldsymbol{J}) + \eta g_\alpha^{(1)}(q_r,q_\theta,\boldsymbol{J}) + \order{\eta^2}, \\
\diff{J_\alpha}{\lambda} = {} & \eta G_\alpha^{(1)}(q_r,q_\theta,\boldsymbol{J}) + \order{\eta^2},
\end{align}
\end{subequations}
where $\lambda$ is Mino time~\cite{Mino2003}, and the forcing functions $g_\alpha^{(1)}$ and $G_A^{(1)}$ originate from the first-order self-force. By working with $\lambda$ instead of proper time $\tau$, the radial and polar motions decouple. At zeroth order in the mass ratio we recover the limit of purely geodesic motion: the integrals of the motion are actually constants and the angle variables evolve according to their associated frequencies $\omega_\alpha$.

@@

One might be concerned about the nature of resonances following the inclusion of the self-force: true geodesic motion only exists at zeroth order in $\eta$ and, while it is a good approximation over short timescales, for small $\eta$ there is a small disparity. The conservative piece of the self-force induces extra precession which leads to a slight shift in the orbital frequencies~\cite{Warburton2012}.\footnote{The Kolmogorov--Arnold--Moser (KAM) theorem states that when an integrable Hamiltonian (i.e. the case for motion in Kerr) is subject to a small perturbation the form of the orbits is preserved albeit slightly deformed~\cite{Arnold1963,Moser1973}. % Chapter II, section 3.3 d)
This should ensure that, in general, there are only small shifts in the orbital frequencies. However, the KAM theory is only valid for sufficiently incommensurate orbits: close to resonance it does not apply~\cite{Moser1973}. % Chapter V section 1. c) 
This is a further reason why resonances merit an in-depth investigation.}
The dissipative piece causes the frequencies to evolve and, hence, the resonance cannot persist for multiple orbits (without some feedback coupling). In effect, we are really considering a period of time about the resonant crossing. The instantaneous orbital frequencies oscillate back and forth around their averaged values. However, there is a time span when the frequencies are consistently close to being commensurate. During this time, the trajectory appears similar to a resonant trajectory, filling only a smaller region of the parameter space. It is this time period that is of interest for transient resonances~\cite{Bosley1992}.

@@

To follow the evolution of the inspiral we must have a means of prescribing the forcing acceleration. We work directly with the gravitational self-force, using the same post-Newtonian (PN) approximation as Flanagan and Hinderer~\cite{Flanagan2012}. For comparison, Flanagan, Hughes and Ruangeri~\cite{Flanagan2012a} use a Teukolsky-equation calculation of GW fluxes to account for the inspiral due to radiation reaction.

The self-force model uses the first-order PN terms of the dissipative self-force formulated in Flanagan and Hinderer~\cite{Flanagan2007} and the conservative force formulated in Iyer and Will~\cite{Iyer1993}, and Kidder~\cite{Kidder1995}. Since only the first PN terms are used, this prescription is of limited validity in strong fields. Both pieces of the self-force are computed assuming that the spin is small: the dissipative piece contains terms to $\order{a_\ast^2}$ and the conservative piece to $\order{a_\ast}$. This is suboptimal for high spins. We also find that this particular implementation of the self-force model marginally overestimates the adiabatic inspiral rate with respect to direct PN evolutions by a factor of $\mathcal{O}(1)$, even for systems in the weak field and with low values of the spin. While this approximate self-force is not perfect, it should serve as a guide for the behaviour of the full self-force, allowing us to assess the qualitative impact of resonances on EMRI detection.

@@

We now study a system that does pass through a resonance during its $2~\mathrm{yr}$ evolution. Specifically, we choose the initial conditions to be the same as in \secref{nonres}, but with an initial semilatus rectum $p_0 = 7.85$. This system passes through the $2$:$3$ resonance; the effect is to cause a shift in the orbital parameters (and hence the fundamental frequencies) that is not replicated by the adiabatic evolution, thus resulting in a rapid dephasing of the waveforms.

@@

\begin{figure}
\centering
\includegraphics[width=0.46\textwidth]{Fig_res_jump_calc}
\caption{\label{fig:res-jump-calc}The difference in energy between the instantaneous model and adiabatic model that matches at the $2$:$3$ resonance, scaled by the magnitude of the $2$:$3$ resonance jump. The dashed green (yellow) lines show the bounding fits to the data before (after) the resonance, used to numerically estimate the size of the jump. The dotted red line indicates the computed size of the jump. The time axis is centred on the $2$:$3$ resonance and is scaled by the resonance timescale $\tau\sub{res}$.}
\end{figure}

@@

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum as a function of eccentricity for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentircity~\cite{ColeThesis2015}. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.

>>>>

\newcommand{\pcm}[1]{{\color{blue}~\textsf{[P: #1]}}}
\newcommand{\cm}[2]{{\color{red}~\textsf{[CITE ME: #2]}}}

@@

\begin{abstract}
The inspiral of stellar-mass compact objects into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect and analyse these signals, accurate waveform templates are required. For computational efficiency, adiabatic templates are often used, which accurately reproduce orbit-averaged trajectories arising from the first-order dissipative gravitational self-force. Other effects are neglected, in particular that of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact object's trajectory and, ultimately, the phase of the emitted radiation. We investigate the properties of transient resonances. We present an evolution scheme for studying the effects of resonances and apply this to an astrophysically motivated population. We find that a large proportion encounter a low-order resonance in the later stages of inspiral; however, the resulting effect on signal-to-noise recovery is small as a consequence of the low eccentricity of the inspirals . Neglecting the effects of transient resonances leads to a loss of $4\%$ of detectable signals.

\pcm{The inspiral of stellar-mass compact objects, like neutron stars or stellar mass black holes, into supermassive black holes provides a wealth of information about the strong gravitational-field regime via the emission of gravitational waves. In order to detect and analyse these signals, waveform templates describing the adiabatic inspiral of the compact star via self-force effects are required. However, other effects are neglected, in particular that of transient resonances, where the radial and poloidal fundamental frequencies become commensurate. During such resonances, the flux of gravitational waves can be diminished or enhanced, leading to a shift in the compact object's trajectory and, ultimately, the phase of the waveform. We present an evolution scheme for studying the effects of transient resonances and apply this to an astrophysically motivated population. We find that a large proportion of systems encounter a low-order resonance in the later stages of inspiral; however, the resulting effect on signal-to-noise recovery is small as a consequence of the low eccentricity of the inspirals. Neglecting the effects of transient resonances leads to a loss of $4\%$ of detectable signals.
}
\end{abstract}

@@

\section{Introduction}
\pcm{The paper is very nice and the result is very important bcs it closes and open topic about the possible bias introduced by resonances in EMRI PE. General comments: (1)check acronyms along the text, in particular gravitational wave and GW, EMRI, etc....!!, (2) we don't mention anything about the astrophysical implications in the intro. It would be nice to include a couple of sentences. (3) I'll emphasise more the final result in the conclusions. }

@@

\pcm{to put things in a historical context, maybe we could include something like this somewhere, : The first detection of GWs represents a paradigm shift in GW astronomy, revolutionising astrophysics and theoretical physics  and changing our understanding of the Universe, perhaps in an even more dramatic way than happened when the X-ray, radio or infrared electromagnetic bands were opened.}

We have made progress in understanding the general relativistic two-body problem in recent years. Bodies of comparable mass can be studied using numerical relativity. Rapid advances in this field have been made following breakthroughs in 2005~\cite{Pretorius2005,Campanelli2006,Baker2006}; it is now possible to simulate hundreds of orbits~\cite{Szilagyi2015}. \pcm{what about EOB+NR??? it's playing an important role...} These simulations allow us to understand binary BH coalescences. Stellar-mass BH mergers are targets for ground-based GW detectors, such as Advanced LIGO~\cite{Aasi2015} and Advanced Virgo~\cite{Acernese2015}, and the in-construction KAGRA~\cite{Aso2013} \pcm{ET...}. The first direct observations of GWs came from the coalescences of two stellar-mass BHs~\cite{Abbott2016,Abbott2016e,Abbott2016d}, and analysis of their properties relied upon our understanding of binary BH waveforms~\cite{Abbott2016f,Abbott2016h,Abbott2016d}. %Supermassive BH (SMBH) coalescences, expected to be the result of galaxy mergers, are potential sources for a space-borne detector such as eLISA~\cite{Amaro-Seoane2012a}.

\pcm{We have made progress in understanding the general relativistic two-body problem in recent years. Bodies of comparable mass can be studied using numerical relativity. Rapid advances in this field have been made following breakthroughs in 2005~\cite{Pretorius2005,Campanelli2006,Baker2006}. The length of these waveforms is limited by their high computational cost, and to generate sufficiently many NR waveforms, long enough to be used directly in GW analysis, is still work in progress ~\cite{Szilagyi2015}. Reliable low-frequency (of the order v/c, where v is the velocity of the system) inspiral waveforms can be obtained using the post-Newtonian (PN) formalism. But PN waveforms are valid for slow motion and hence they are not valid in the strong field regime close to merger. For this reason, the effective-one-body (EOB)formalism, a hybrid of PN and NR, is gaining popularity. On the one hand, it provides a complete description of the coalescence of compact object binaries and, on the other hand, it is computationally cheaper compared to full NR simulations. EOB produces waveforms for the inspiral, merger and ringdown stages using PN and black-hole perturbation theory results. Hence, the EOB model reduces to the PN approximation at low frequency. In the strong- field regime, however, EOB models are calibrated to existing NR simulations. These waveforms allow us to understand binary BH coalescences....}

@@

In the case of extreme-mass-ratio systems, efforts are concentrated on understanding the gravitational self-force~\cite{Barack2009,Poisson2004}\pcm{need up-to-date references, like Isoyama et al 2014,  Pound 2012, Hughes et al 2015}. In the test-particle limit, the smaller body follows an exact geodesic of the SMBH's spacetime. Including the effects of the smaller body's finite mass, the background spacetime is perturbed. The back-reaction from this deformation alters the small body's orbital trajectory, and can be modelled as a self-force that moves the body from its geodesic. The self-force is commonly divided into two pieces, dissipative and conservative \pcm{cites???}. The former encapsulates the slow decay of the orbital energy and angular momentum, constants of the motion in the test particle limit, through radiation. The latter shifts the orbital phases inducing precession. The dissipative piece is time-asymmetric and has the larger effect on the evolution of the orbital phase; the conservative piece is time-symmetric and has a smaller influence on the phase, although this can accumulate over many orbits. Being able to accurately model the influence of the self-force allows us to create reliable waveform models.

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$.\footnote{In the strong-field regime, it is possible to have isofrequency pairings, where two different orbits share the same orbital frequencies~\cite{Warburton2013}. The evolution of the frequencies still differ, such that orbital trajectories can be reconstructed.} The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field regime $\Omega_r < \Omega_\theta < \Omega_\phi$, and they may differ significantly. For EMRIs, the evolution \pcm{inspiral} timescale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital timescales. The evolution  \pcm{inspiral}  of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate\pcm{, and resonances occur...}. Resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu \equiv \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

Resonances involving the azimuthal motion do not produce a comparable effect because of the axisymmetry of the background spacetime. However, both $\theta$--$\phi$ resonances~\cite{Hirata2011} and $r$--$\phi$ resonances~\cite{VanDeMeent2013} can lead to extrinsic effects; the GWs from such systems are not emitted isotropically and the imbalance produces a kick velocity that is, in some cases, sufficient to eject the central BH from its host \pcm{cite??}.

Geodesic motion in Kerr spacetime can be described by use of the action--angle formalism~\cite{Goldstein2002}. % Chapter 10
We consider a body of mass $\mu$ orbiting a BH of mass $M$, with $\eta = \mu/M \ll 1$,\footnote{To first order, the mass ratio $\eta$ is the same as the symmetric mass ratio $\mu M/(\mu+M)^2$.} and describe the motion in the directions of the standard Boyer--Lindquist coordinates $\{t,r,\theta,\phi\}$ \pcm{cite} using generalised angle variables $q_\alpha = \{q_t,q_r,q_\theta,q_\phi\}$ \citep{Hinderer2008}. We denote the first integrals of the geodesic motion, the generalised action variables, by $J_\alpha$. These are some combination of the energy per unit mass $E$ and the axial angular momentum per unit mass $L_z$ \pcm{of the orbit}, which arise from isometries of the metric in $t$ and $\phi$, and the Carter constant per unit mass squared $Q$~\cite{Carter1968}, which is related to the separability of the equations of motion in $r$ and $\theta$. The system evolves following~\cite{Flanagan2012}
\begin{subequations}
\label{eq:Mino-E-o-M}
\begin{align}
\diff{q_\alpha}{\lambda} = {} & \omega_\alpha(\boldsymbol{J}) + \eta g_\alpha^{(1)}(q_r,q_\theta,\boldsymbol{J}) + \order{\eta^2}, \\
\diff{J_\alpha}{\lambda} = {} & \eta G_\alpha^{(1)}(q_r,q_\theta,\boldsymbol{J}) + \order{\eta^2},
\end{align}
\end{subequations}
where $\lambda$ is Mino time~\cite{Mino2003} \pcm{not everyone is familiar with this term. Besides, Mino does't like it. We could consider calling it "effective time" or similar}, and the forcing functions $g_\alpha^{(1)}$ and $G_A^{(1)}$ originate from the first-order self-force \pcm{this implies that there are  second, etc orders, we need to mention it and cite appropriately}. By working with $\lambda$ instead of proper time $\tau$, the radial and polar motions decouple. At zeroth order in the mass ratio we recover the limit of purely geodesic motion: the integrals of the motion are actually constants and the angle variables evolve according to their associated frequencies $\omega_\alpha$.

@@

\pcm{we are including a lot of material that can be found elsewhere and that we are not actually using directly. This makes the paper unecessary long. We could consider reducing the prevous part or including it in the appendix.}
One might be concerned about the nature of resonances following the inclusion of the self-force: true geodesic motion only exists at zeroth order in $\eta$ and, while it is a good approximation over short timescales, for small $\eta$ there is a small disparity. The conservative piece of the self-force induces extra precession which leads to a slight shift in the orbital frequencies~\cite{Warburton2012}.\footnote{The Kolmogorov--Arnold--Moser (KAM) theorem states that when an integrable Hamiltonian (i.e. the case for motion in Kerr) is subject to a small perturbation the form of the orbits is preserved albeit slightly deformed~\cite{Arnold1963,Moser1973}. % Chapter II, section 3.3 d)
This should ensure that, in general, there are only small shifts in the orbital frequencies. However, the KAM theory is only valid for sufficiently incommensurate orbits: close to resonance it does not apply~\cite{Moser1973}. % Chapter V section 1. c) 
This is a further reason why resonances merit an in-depth investigation.}
The dissipative piece causes the frequencies to evolve and, hence, the resonance cannot persist for multiple orbits (without some feedback coupling). In effect, we are really considering a period of time about the resonant crossing. The instantaneous orbital frequencies oscillate back and forth around their averaged values. However, there is a time span when the frequencies are consistently close to being commensurate. During this time, the trajectory appears similar to a resonant trajectory, filling only a smaller region of the parameter space. It is this time period that is of interest for transient resonances~\cite{Bosley1992}.

@@

To follow the evolution of the inspiral we must have a means of prescribing the forcing acceleration \pcm{that deviates the orbit from a geodesic motion}. We work directly with the gravitational self-force, using the same post-Newtonian (PN) approximation as Flanagan and Hinderer~\cite{Flanagan2012}. For comparison, Flanagan, Hughes and Ruangeri~\cite{Flanagan2012a} use a Teukolsky-equation calculation of GW fluxes to account for the inspiral due to radiation reaction.

The self-force model uses the first-order PN terms of the dissipative self-force formulated in Flanagan and Hinderer~\cite{Flanagan2007} and the conservative force formulated in Iyer and Will~\cite{Iyer1993}, and Kidder~\cite{Kidder1995}. Since only the first PN terms are used, this prescription is of limited validity in strong fields. Both pieces of the self-force are computed assuming that the \pcm{BH} spin is small: the dissipative piece contains terms to $\order{a_\ast^2}$ and the conservative piece to $\order{a_\ast}$. This is suboptimal for high spins. We also find that this particular implementation of the self-force model marginally overestimates the adiabatic inspiral rate with respect to direct PN evolutions by a factor of $\mathcal{O}(1)$, even for systems in the weak field and with low values of the spin. While this approximate self-force is not perfect, it should serve as a guide for the behaviour of the full self-force, allowing us to assess the qualitative impact of resonances on EMRI detection.

@@

We now study a system that does pass through a resonance during its $2~\mathrm{yr}$ evolution. Specifically, we choose the initial conditions to be the same as in \secref{nonres}, but with an initial semilatus rectum $p_0 = 7.85$. This system passes through the $2$:$3$ resonance; the effect is to cause a shift in the orbital parameters (and hence the fundamental frequencies) that is not replicated by the adiabatic evolution, thus resulting in a rapid dephasing of the waveforms. \pcm{perhaps we could mention why we choose this one in particular, i.e. it's the strongest one that we found from the start of the evolutions at p =xxx}

@@

\begin{figure}
\centering
\includegraphics[width=0.46\textwidth]{Fig_res_jump_calc}
\caption{\label{fig:res-jump-calc}The difference in energy between the instantaneous model and adiabatic model that matches at the $2$:$3$ resonance, scaled by the magnitude of the $2$:$3$ resonance jump. The dashed green (yellow) lines show the bounding fits to the data before (after) the resonance, used to numerically estimate the size of the jump. The dotted red line indicates the computed size of the jump. The time axis is centred on the $2$:$3$ resonance and is scaled by the resonance timescale $\tau\sub{res}$. \pcm{plots too small}}
\end{figure}

@@

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum \pcm{p} as a function of eccentricity \pcm{e} for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentircity~\cite{ColeThesis2015}. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.
